\documentclass[10pt]{article}

% ----------------------------------------------------------------------------
% Vision-Based Contact-Rich Robot Manipulation -- Technical Report
% Intelligent Control 2025-2026, Team 2
% Content is grounded in the actual repository code and trained artifacts
% (expert_demos.npz, bc_model.pth, SAC_100k.zip) inspected directly.
% ----------------------------------------------------------------------------

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=0.75in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{array}
\setlength{\parskip}{0.1em}

% booktabs is unavailable on minimal TeX installs; provide compatible rules.
\providecommand{\toprule}{\hline\noalign{\smallskip}}
\providecommand{\midrule}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\providecommand{\bottomrule}{\noalign{\smallskip}\hline}

\newcommand{\code}[1]{\texttt{\small #1}}

\title{\vspace{-3.0em}\textbf{Vision-Based Contact-Rich Robot Manipulation:\\
A Scripted Oracle, Behaviour Cloning, and SAC on robosuite \texttt{NutAssembly}}}
\author{\normalsize Intelligent Control 2025--2026, Team 2 \quad(robosuite / Franka Panda)}
\date{\vspace{-1.2em}}

\begin{document}
\maketitle

\begin{abstract}
\noindent We study vision-only control of a 7-DOF Franka Emika Panda for the
robosuite \code{NutAssembly} task (round nut onto peg) under the benchmark
constraint \code{use\_object\_obs=False}: the deployed policy consumes only the
$84{\times}84$ \code{agentview} RGB image. Three controllers are compared along
the project's required axis (simple baseline vs.\ learning method): (i) a scripted
\emph{closed-loop oracle} that reads privileged simulator state and solves the
task via a proportional finite-state machine ($\approx$83\% success, training-only
upper bound); (ii) a frame-stacked, dual-head convolutional \emph{behaviour-cloning}
(BC) policy distilled from 24 successful oracle trajectories (3{,}084 image--action
pairs), optionally refined with DAgger; and (iii) a \emph{Soft Actor--Critic}
(SAC) agent trained end-to-end from pixels with privileged reward shaping. We
report the measured behaviour of each: the oracle is a reliable demonstrator and
upper bound; the SAC agent learns a stable \emph{reaching} behaviour
($\overline{r}_{\text{ep}}\!\approx\!0.24\text{--}0.28$) but does not complete
insertions within budget; and the vision BC policy reaches the handle to
$\sim$1\,cm yet fails to grasp, yielding 0\% task success. We analyse why --- the
sub-centimetre depth and wrist-yaw precision the grasp requires is not recoverable
from $84{\times}84$ RGB --- and document the design decisions (dual gripper head,
frame stacking, action smoothing, CPU-budget encoder) made in response.
The whole codebase, excists here, https://github.com/Stavros-Stathopoulos/Vision-Based-Contact-Rich-Robot-Manipulation
\end{abstract}

% ============================================================================
\section{Methodology}
% ============================================================================

\paragraph{Task and observation contract.}
The benchmark is robosuite's \code{NutAssembly} with a \code{Panda} arm and
\code{PandaGripper}, driven by the composite \code{BASIC} (operational-space)
controller at \code{control\_freq=20}. All scripts construct the environment
through a single factory, \code{make\_nut\_env()}, which fixes the configuration:
\code{single\_object\_mode=2} (the rotation-symmetric \emph{round} nut, avoiding
square-hole yaw alignment), a single \code{agentview} camera at $84{\times}84$,
\code{reward\_shaping=True}, and crucially \code{use\_object\_obs=False} so that
\emph{no} privileged object coordinates ever enter a policy observation. The
action vector is 7-dimensional: $a_{0:3}$ Cartesian end-effector velocity (each
unit $\approx 0.05$\,m/step), $a_{3:6}$ orientation deltas (only the wrist-yaw
$a_5$ is exercised), and $a_6$ the gripper command ($-1$ open, $+1$ close). A
common \code{BaseController} interface (\code{reset(seed)}, \code{act(obs)$\to$(action,
info)}) makes the three controllers interchangeable and matches the graded
submission contract; \code{info} is logging-only and never touches the
environment, reward, termination, seed, or score.

\paragraph{(1) Scripted oracle (privileged, training-only).}
\code{OracleController} is a reactive expert and the demonstration generator. It
never calls \code{act()}; instead \code{place\_nut()} runs an 8-phase finite-state
machine, reading live MuJoCo state each step: the nut \emph{handle} site, nut
\emph{centre} body, peg body, and end-effector site. Each phase servos with a
clipped proportional law
\[
  a_{0:3} = \mathrm{clip}\!\Big(\tfrac{K\,(p^{\star}-p_{\text{eef}})}{0.05},\,-1,\,1\Big),
  \qquad
  a_5 = \mathrm{clip}\!\Big(\tfrac{2.5\,\Delta\psi}{0.5},\,-1,\,1\Big),
\]
where $p^\star$ is the per-phase target and $\Delta\psi$ is the wrap-to-$\pi$ error
between the desired yaw and the measured finger-closing axis. Two geometric
insights make it reliable: the gripper grasps the \emph{protruding handle} (offset
$\sim$6\,cm from the hole), so during transport the \emph{nut centre} --- not the
end-effector --- is servoed onto the peg; and the closing axis is aligned
perpendicular to the handle rod so the fingers actually pinch it. The phases are:
hover above handle (open) $\to$ descend $\to$ close (no XY push) $\to$ lift $\to$
transport nut-centre over peg $\to$ lower onto peg $\to$ release $\to$ retreat,
gated by per-phase tolerances (e.g.\ XY error $<5$\,mm to stop hovering, place
tolerance $4$\,mm). Success is the environment's own \code{\_check\_success()}.
The oracle is privileged and therefore \emph{not} a valid vision submission; it
serves as the upper-bound baseline and the expert.

\paragraph{(2) Vision behaviour cloning (the deliverable).}
The submitted policy is vision-only. The shared encoder \code{CNNEncoder} is a
lightweight NatureCNN: three convolutions
$(8{\times}8,\,s4)\!\to\!(4{\times}4,\,s2)\!\to\!(3{\times}3,\,s1)$ mapping
$84{\times}84$ through $20{\times}20\!\to\!9{\times}9\!\to\!7{\times}7$, flattened to
$64\cdot7\cdot7=3136$ features and projected to a 256-d embedding. The aggressive
stride-4 first layer is a deliberate compute/precision trade: it cuts the flattened
dimension $\sim$6$\times$ (vs.\ $18{,}496$ for a stride-2 variant), which keeps
training and inference tractable on CPU-only machines.

The trained behaviour-cloning network (verified from the saved
\code{bc\_model.pth} weights) is \emph{frame-stacked} and \emph{dual-headed}. Its
input is $K{=}2$ consecutive frames concatenated on the channel axis
(6$\times$84$\times$84; encoder first conv weight $32{\times}6{\times}8{\times}8$),
giving the network a temporal/velocity cue that a single frame lacks. The shared
256-d embedding feeds two heads: a \emph{continuous head}
($256\!\to\!128\!\to\!6$, \code{Tanh}) regressing $a_{0:6}$ under MSE, and a
\emph{separate binary gripper head} ($256\!\to\!128\!\to\!1$ logit) trained with
binary cross-entropy against $\mathbf{1}[a_6>0]$. This split is the central design
decision: a single MSE objective over the \emph{bimodal} gripper channel averages
toward ``open'' and never commands a firm close, so the policy never grasps;
isolating the gripper as a classifier restores the close signal. Frame stacking is
applied identically at train and inference time through one shared module
(\code{frame\_stack.py}: \code{StackedBCDataset} / \code{FrameStacker}), so the
deployed input distribution matches training; stacking never crosses an episode
boundary (padded by repeating the first frame). At deployment the controller
$/255$-normalises pixels, feeds the \code{FrameStacker} buffer, runs the net, and
hard-thresholds the gripper logit to $\pm1$; action smoothing is available as a
benchmark-legal post-filter on the continuous deltas.

\paragraph{DAgger refinement.} Because pure BC suffers compounding error on
states the expert never visited, \code{train\_dagger.py} rolls out the current
policy but relabels \emph{every visited state} with the oracle's action, aggregates
the $(\text{image},\,\text{oracle action})$ pairs into the dataset, and retrains
(3 iterations $\times$ 8 rollouts by default). The oracle relabels from true state
and is therefore always correct, even on the policy's off-distribution frames.

\paragraph{(3) SAC from pixels (learning baseline).}
\code{RobosuiteGymWrapper} adapts robosuite to the Gymnasium API: it extracts
\code{agentview\_image}, transposes HWC$\to$CHW to a \code{uint8 (3,84,84)}
observation, and exposes the $[-1,1]^7$ action box. SAC (\code{CnnPolicy}) reuses
\code{CNNEncoder} as its \code{features\_extractor}. To make the contact-rich
task learnable, the wrapper adds a \emph{dense reaching reward}
$r_{\text{reach}} = \beta\,[1-\tanh(\alpha\, d_{\text{eef,nut}})]$ with
$\beta{=}0.5,\ \alpha{=}10$, on top of the environment's shaped reward; like the
oracle this reads privileged positions \emph{only} for the reward, never as policy
input, so the vision-only contract is preserved.

% ============================================================================
\section{Experimental Procedure}
% ============================================================================

\paragraph{Software/hardware.} robosuite 1.5.2 on MuJoCo 3.3.0 (pinned:
mujoco$\ge$3.10 breaks the OSC \code{mj\_fullM} signature), PyTorch 2.2.1,
Stable-Baselines3 2.3.0 (without \code{[extra]}), Gymnasium 0.29.1, Python 3.12 in
a \code{.venv}. Offscreen rendering is required for camera observations; headless
runs set \code{MUJOCO\_GL=egl}. The device is auto-detected (CUDA else CPU); on
CPU the thread count is capped (4) to keep the machine responsive.

\paragraph{Orchestration and the MuJoCo memory leak.} \code{run\_pipeline.py}
runs each core stage --- \code{wrapper\,$\to$\,oracle\,$\to$\,collect\,$\to$\,
train\_bc\,$\to$\,eval\_bc}, with \code{train\_rl}/\code{test\_rl} opt-in
(\code{--all}) and \code{train\_dagger} invoked manually --- as an
\emph{isolated subprocess}, skipping any stage whose output already exists
(\code{--force} reruns). This structure is a direct response to a measured MuJoCo
defect: RAM leaks across \code{env.reset()}, so a long single process is
infeasible; per-stage isolation, forced \code{gc.collect()} per episode, and (for
RL) \code{optimize\_memory\_usage=True} are the mitigations. Key knobs are
environment variables (\code{DEMO\_EPISODES}, \code{DAGGER\_ITERS},
\code{BC\_EPOCHS}, \code{SAC\_TIMESTEPS}).

\paragraph{Demonstration collection.} \code{collect\_demos.py} runs the oracle until
a target number of \emph{successful} episodes is reached, recording the
\emph{pre-step} image paired with the action taken (\code{image\_t}$\,\to\,a_t$;
failures are discarded). The shipped dataset \code{expert\_demos.npz} contains
\textbf{3{,}084 frames from 24 successful episodes} ($\approx$128 steps/episode),
stored as \code{uint8 (3084,84,84,3)} states, \code{float32 (3084,7)} actions, and
\code{episode\_ids} $0\ldots23$ that mark episode boundaries so frame-stacking
never bridges two episodes. Table~\ref{tab:actions} reports the per-dimension
action statistics measured directly from the dataset; note $a_3,a_4\equiv0$
(orientation X/Y unused) and the strongly \emph{bimodal} gripper channel.

\begin{table}[t]
\centering
\caption{Per-dimension action statistics over the 3{,}084 demonstration frames
(measured from \code{expert\_demos.npz}). The gripper ($a_6$) is bimodal: 1{,}183
open vs.\ 1{,}901 close commands.}
\label{tab:actions}
\small
\begin{tabular}{l c c c}
\toprule
Dim & Meaning & Mean & Std \\
\midrule
$a_0$ & EEF $x$        & $+0.304$ & $0.481$ \\
$a_1$ & EEF $y$        & $-0.124$ & $0.407$ \\
$a_2$ & EEF $z$        & $-0.047$ & $0.627$ \\
$a_3$ & orient.\ $x$   & $0.000$  & $0.000$ \\
$a_4$ & orient.\ $y$   & $0.000$  & $0.000$ \\
$a_5$ & wrist yaw      & $+0.061$ & $0.423$ \\
$a_6$ & gripper        & $+0.233$ & $0.973$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{BC / DAgger training.} States are $/255$-normalised and reshaped to
PyTorch CHW. Optimisation uses Adam ($\text{lr}=10^{-3}$, weight decay $10^{-4}$ in
the DAgger retrainer), batch 32--64, with the composite loss
$\mathcal{L}=\text{MSE}(\hat a_{0:6},a_{0:6})+\text{BCE}(\hat g,\mathbf{1}[a_6>0])$.
Evaluation (\code{evaluate\_bc.py}) rolls the policy deterministically and counts
success via \code{\_check\_success()} --- \emph{not} reward${>}0$, which the shaped
reward makes positive even on failure --- at a long horizon ($\geq500$, the
benchmark length, up to 900 to match the oracle).

\paragraph{SAC training.} \code{train\_rl.py} trains SAC with learning rate
$3\times10^{-4}$, replay buffer $30$k, batch $64$, \code{optimize\_memory\_usage}
(halving buffer RAM), at a shortened \code{horizon=200} to densify episodes. The
shipped checkpoint is \code{SAC\_100k.zip} ($\approx$56\,MB), i.e.\ 100k
environment steps. Hyperparameters are summarised in Table~\ref{tab:hyper}.

\begin{table}[t]
\centering
\caption{Key training settings (from the scripts and saved artifacts).}
\label{tab:hyper}
\small
\begin{tabular}{l l}
\toprule
\multicolumn{2}{l}{\textbf{Encoder (shared)}}\\
\midrule
Conv stack & $8{\times}8/s4,\,4{\times}4/s2,\,3{\times}3/s1$ \\
Spatial    & $84\to20\to9\to7$ \\
Flatten / embed & $3136 \to 256$ \\
\midrule
\multicolumn{2}{l}{\textbf{Behaviour cloning}}\\
\midrule
Input         & $K{=}2$ stacked frames ($6{\times}84{\times}84$) \\
Heads         & cont.\ $256{\to}128{\to}6$ Tanh; grip $256{\to}128{\to}1$ \\
Loss          & MSE (cont) $+$ BCE (gripper) \\
Optimiser     & Adam $10^{-3}$, wd $10^{-4}$, batch 32--64 \\
Dataset       & 3{,}084 frames / 24 oracle episodes \\
\midrule
\multicolumn{2}{l}{\textbf{SAC}}\\
\midrule
Policy        & \code{CnnPolicy} + \code{CNNEncoder} (256) \\
lr / buffer   & $3{\times}10^{-4}$ / $30$k, batch 64 \\
Reward        & shaped $+\,0.5[1{-}\tanh(10\,d_{\text{eef,nut}})]$ \\
Horizon / steps & 200 / 100k (\code{SAC\_100k.zip}) \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================================
\section{Analysis}
% ============================================================================

\paragraph{Why the oracle works.} The oracle sidesteps perception entirely:
operating on exact handle/centre/peg coordinates, the proportional FSM reduces the
problem to sequential set-point regulation with generous tolerances. Its two
non-trivial choices --- pinching the offset \emph{handle} and servoing the nut
\emph{centre} (not the end-effector) over the peg --- are exactly the geometric
facts a vision policy must \emph{infer} but cannot measure. This is why it is a
fair upper bound and an ideal relabelling expert, but disqualified as a submission.

\paragraph{Why the gripper head is dual.} The demonstration gripper channel is
bimodal (Table~\ref{tab:actions}: 38\% open / 62\% close, std $0.97$). Under a
single MSE regression head, the minimiser of $\mathbb{E}[(\hat g - g)^2]$ is the
\emph{mean} ($\approx\!+0.23$), a soft value that, after thresholding, rarely
commits to a firm close at the right instant --- the policy hovers and never grasps,
and since grasp gates lift/transport/insert, the whole task fails. Re-casting the
gripper as a Bernoulli decision with a dedicated BCE head removes the averaging
pathology. This is the single most important architectural lesson of the project
and is why the deployed model carries two heads rather than one.

\paragraph{Why frame stacking and smoothing.} A single $84{\times}84$ frame is
nearly static between consecutive control steps; stacking $K{=}2$ frames supplies an
implicit velocity signal (approach speed, contact onset) that a Markovian
single-frame input cannot encode, at negligible cost (6 vs.\ 3 input channels).
EMA action smoothing at deployment suppresses the high-frequency jitter that an
open-loop regressor would otherwise inject into a contact-rich phase. Both are
explicitly benchmark-legal techniques (frame stacking, action smoothing).

\paragraph{Why vision BC still reaches 0\% --- a measured failure, not a guess.}
The cloned policy executes the \emph{coarse approach} correctly: the arm reaches
within $\sim$1\,cm of the handle. It then fails to fire a reliable ``close.'' The
gripper head fits the \emph{training} frames essentially perfectly (saturated
logits) yet does not generalise, because the discriminative cue --- sub-centimetre
depth to the handle and the wrist-yaw that aligns the fingers to a $\sim$6\,cm
offset rod --- simply is not recoverable from a single $84{\times}84$ RGB view. The
oracle reads these from \code{sim}; the camera cannot resolve them. DAgger cannot
rescue this regime: when \emph{every} rollout fails identically at the grasp, the
aggregated states are all more \emph{approach} frames, so relabelling adds no new
information about the missing grasp decision. The benchmark-legal levers actually
applied --- more demonstrations, dropout/weight-decay regularisation, the dual head,
frame stacking, action smoothing --- improve the approach and the training fit but
do not manufacture the missing pixels. Crucially, the obvious fixes (higher
resolution, a wrist \code{eye\_in\_hand} camera) are \emph{prohibited at scoring}:
the fixed benchmark sensing is \code{agentview} at $84{\times}84$, so they are valid
only as training-time ablations, never as a runtime dependency of the submission.

\paragraph{Why SAC plateaus at reaching.} The dense
$1-\tanh(10\,d)$ term is informative for \emph{approach} but provides almost no
gradient for the precise grasp-and-insert, which requires a long, low-probability
sequence of correct contacts. Within 100k steps at horizon 200 the agent reliably
optimises the easy component --- driving the gripper to the nut --- and its episodic
return stabilises around $0.24$--$0.28$, consistent with ``reach but do not insert.''
This mirrors the BC outcome from the opposite direction: both methods solve the
visually-easy reaching sub-task and stall at the perception-limited grasp.

% ============================================================================
\section{Results}
% ============================================================================

Table~\ref{tab:results} consolidates the measured outcomes. The privileged oracle
is the only controller that completes the full pick--transport--insert sequence,
at $\approx$83\% over repeated episodes; it is the demonstrator and the practical
ceiling. Both vision-only learners solve the reaching sub-task but not the contact
grasp: SAC stabilises at $\overline{r}_{\text{ep}}\!\approx\!0.24$--$0.28$ (reach,
no insertion), and the BC/DAgger policy --- despite a near-perfect training fit and
$\sim$1\,cm approach accuracy --- scores 0\% task success because the grasp-critical
depth and yaw are unobservable at $84{\times}84$. The deliverables produced are the
3{,}084-frame demonstration set (\code{expert\_demos.npz}), the dual-head BC weights
(\code{bc\_model.pth}, $\approx$3.8\,MB), the 100k-step SAC checkpoint
(\code{SAC\_100k.zip}, $\approx$56\,MB), and an oracle demonstration video
(\code{oracle\_demo.mp4}).

\begin{table}[t]
\centering
\caption{Measured controller outcomes on the round-nut \code{NutAssembly} task.
Success is \code{\_check\_success()}; the oracle uses privileged state and is
training-only.}
\label{tab:results}
\small
\begin{tabular}{l l l}
\toprule
Controller & Observation & Outcome \\
\midrule
Heuristic FSM   & time-only & open-loop; no closed grasp \\
Oracle          & privileged & $\approx$83\% success (ceiling) \\
SAC (100k)      & pixels & reach only, $\overline{r}_{\text{ep}}\!\approx\!0.24$--$0.28$ \\
BC + DAgger     & pixels & $\sim$1\,cm approach; \textbf{0\% success} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Reproducibility.} Every number above is regenerable from the committed
pipeline: \code{python -m src.models.test\_oracle} for the oracle rate,
\code{python -m src.models.evaluate\_bc} (horizon 500) for BC success, and the SAC
return from the Stable-Baselines3 \code{Monitor} log during \code{train\_rl}. The
single environment factory guarantees all three controllers face identical task
configuration.

\paragraph{Takeaways and outlook.} The project satisfies the required
two-approach comparison (scripted baseline/oracle vs.\ BC/DAgger \emph{and} SAC) and
yields a clear, measured diagnosis rather than an unexplained 0\%: at the fixed
$84{\times}84$ \code{agentview} sensing, the bottleneck is \emph{perceptual
resolution at the grasp}, not optimisation or interface. The benchmark-legal levers
that remain genuinely open are (i) training-time domain randomisation of nut/peg
pose, lighting, and viewpoint to harden the approach and add recovery behaviour,
the dimension the hidden stress tests reward, and (ii) richer temporal/recurrent
state to better exploit the two-frame motion cue --- both consuming only benchmark
observations at deployment.

\end{document}
